---
title: "soce_dzyn26_initial_primer_design"
format: html
---

```{r}
library(tidyverse)
library(readxl)
library(here)
```


I followed Tom Delomas' primer design pipeline to generate 500 primers for a new GT-seq panel for red drum.

# Raw reads

Chris Hollenbeck provided whole genome sequence data from fish genotyped during his PhD project. They are located on the HPC in /marinegenomics/projects/fish/red_drum/drum_pop_wgs/raw/. Below is a quick overview of their metadata:

```{r}
meta_data <- readxl::read_xlsx(here::here("data","raw","drum_wgs_samples.xlsx"))
```

```{r}
knitr::kable(
  meta_data,
  digits = 2,
  caption = "Metadata of WGS Red Drum Samples"
)
```

# Genotyping

The raw reads were genotyped using Eriq Anderson's mega-non-model-wgs-snakeflow pipeline https://github.com/eriqande/mega-non-model-wgs-snakeflow

There were several adjustments that were required before the pipeline could function normally. Details of adjustments made can be found in another document, "mega_snakeflow_on_crest."

QC of the genotyping can be found here:

[View QC results](../data/raw/multiqc.html)

The working SNP set used for primer design was made from manually extracting biallelic snps from the pass-maf-0.05 bcf file generated from the pipeline. 

The pass-maf-0.05 file was generated by these filters in GATK:

QD < 2.0                  filtered
QUAL < 30.0               filtered
SOR > 3.0                 filtered
FS > 60.0                 filtered
MQ < 40.0                 filtered
MQRankSum < -12.5         filtered
ReadPosRankSum < -8.0     filtered

The working set of SNPs was called red_drum_maf005.snps.vcf.gz and was generated by manually selected biallelic snps with: bcftools view -v snps -m2 -M2

# Phasing with WhatsHap

Each sample was phased independently using WhatsHap. The phased per-sample VCFs were generated from the filtered SNP VCF and corresponding overlap-clipped BAM files.

A representative command was:

whatshap phase \
  --reference /work/marinegenomics/jmatt1/wgs/mega_soce_cmh/resources/genome.fasta \
  --sample SAMPLE_NAME \
  --output SAMPLE.whatshap.vcf.gz \
  red_drum.maf005.snps.vcf.gz \
  overlap_clipped/SAMPLE.bam

Each phased VCF was indexed, and all 20 single-sample phased VCFs were merged into a single multi-sample phased VCF:

bcftools merge single_sample_vcfs/*.whatshap.vcf.gz \
  -Oz -o allWhatshap_red_drum_maf005.vcf.gz

bcftools index -t allWhatshap_red_drum_maf005.vcf.gz

The merged VCF contained 20 samples and phased genotype information.

# Population-specific VCFs

The merged phased VCF was split into WGULF and EGULF population-specific VCFs using sample lists.

bcftools view -S WGULF.samples.txt \
  allWhatshap_red_drum_maf005.vcf.gz \
  -Oz -o allWhatshap_WGULF.vcf.gz

bcftools index -t allWhatshap_WGULF.vcf.gz

bcftools view -S EGULF.samples.txt \
  allWhatshap_red_drum_maf005.vcf.gz \
  -Oz -o allWhatshap_EGULF.vcf.gz

bcftools index -t allWhatshap_EGULF.vcf.gz

Each population-specific VCF contained 10 samples.

# Optinal HWE screening

To mimic the oyster-panel workflow, Hardy-Weinberg Equilibrium filtering was evaluated separately within WGULF and EGULF using PLINK.

VCF variant IDs were first set using chromosome and position:

bcftools annotate --set-id '%CHROM:%POS' allWhatshap_EGULF.vcf.gz -Oz -o allWhatshap_EGULF.ids.vcf.gz
bcftools annotate --set-id '%CHROM:%POS' allWhatshap_WGULF.vcf.gz -Oz -o allWhatshap_WGULF.ids.vcf.gz

PLINK HWE tests were run with:

plink --vcf allWhatshap_EGULF.ids.vcf.gz --double-id --allow-extra-chr --hardy --out EGULF_ids
plink --vcf allWhatshap_WGULF.ids.vcf.gz --double-id --allow-extra-chr --hardy --out WGULF_ids

SNPs with HWE p-value < 0.01 in either population were counted. A total of 70,715 SNPs failed this criterion, representing approximately 1.26% of SNPs tested. Because this was a low proportion of the total SNP set, the primer-design workflow proceeded using the original SNP set.

# Microhaplotype expected heterozygosity

Microhaplotype expected heterozygosity was calculated using CalcHe_mh_vcf.jar.

For each population VCF, two window sizes were evaluated:

50 bp windows, corresponding to the 75 bp sequencing-read design
125 bp windows, corresponding to the 150 bp sequencing-read design

Example commands:

java -jar CalcHe_mh_vcf.jar -v allWhatshap_WGULF.vcf.gz -w 50
mv vcf_he_output.txt He_50_WGULF.txt

java -jar CalcHe_mh_vcf.jar -v allWhatshap_WGULF.vcf.gz -w 125
mv vcf_he_output.txt He_125_WGULF.txt

The same commands were run for EGULF, producing:

He_50_WGULF.txt
He_125_WGULF.txt
He_50_EGULF.txt
He_125_EGULF.txt

# Panel design inputs

Panel design was performed in R using a modified version of the oyster microhaplotype panel-design script.

The population input section was updated to:

inputfiles <- tibble(
  pop = c("WGULF", "EGULF")
)

For 75 bp read designs, the script used the 50 bp He files:

He_50_WGULF.txt
He_50_EGULF.txt

For 150 bp read designs, the script used the 125 bp He files:

He_125_WGULF.txt
He_125_EGULF.txt

Expected heterozygosity was summarized across populations using the minimum value:

summaryFunction <- min

This prioritizes loci that are informative in both WGULF and EGULF rather than loci that are highly informative in only one population.

Other changes of the oyster panel design script were specifying the correct genome:

refGenome <- "/work/marinegenomics/jmatt1/wgs/mega_soce_cmh/resources/genome.fasta"

and using a Bowtie2 index for red drum.

A Bowtie2 index was built from the same reference genome:

bowtie2-build genome.fasta genome_bt2

and was specified in the panel design R script:

bowtie2_ref <- "/work/marinegenomics/jmatt1/wgs/mega_soce_cmh/resources/genome_bt2"

Chromosome names and lengths were also added to the R script. These were taken from the ref genome index file.

# Primer design and filtering strategy

Primer design was performed with Primer3 through the panel-design R script.

The Primer3 executable was provided by the microhap conda environment:

primer3path <- "/home/jmatt1/.conda/envs/microhap/bin/primer3_core"

Primer design was run during each iteration of the greedy selection algorithm. A locus was only selected if valid primers could be designed and the primer pair mapped uniquely to the reference genome using Bowtie2.

The greedy algorithm balanced:

high minimum expected heterozygosity across WGULF and EGULF,
even genomic distribution,
successful Primer3 primer design,
unique primer-pair mapping.

Eight panel-design runs were evaluated:

Rscript panel_design_drum.R red_drum_00_75 0 0 75
Rscript panel_design_drum.R red_drum_10_75 1 0 75
Rscript panel_design_drum.R red_drum_01_75 0 1 75
Rscript panel_design_drum.R red_drum_11_75 1 1 75

Rscript panel_design_drum.R red_drum_00_150 0 0 150
Rscript panel_design_drum.R red_drum_10_150 1 0 150
Rscript panel_design_drum.R red_drum_01_150 0 1 150
Rscript panel_design_drum.R red_drum_11_150 1 1 150

The first numeric argument controlled the no-recombination/complexity filter:

0: allow all loci
1: retain only loci where number of observed haplotypes was no greater than number of SNPs + 1

The second numeric argument controlled adjacent-SNP filtering:

0: allow adjacent SNPs
1: remove loci containing adjacent SNPs

The third numeric argument controlled the intended read-length design:

75: use 50 bp microhap windows
150: use 125 bp microhap windows

# Output files 

Each panel-design run produced:

<prefix>panel.txt
<prefix>primers.txt
<prefix>pos.txt
<prefix>fwd.txt
<prefix>ampRef.fa

To mimic the mimic the oyster panel design criteria, the 75 bp panels are most appropriate. To target a well-behaving panel, the recombination filter and adjacent SNP-filtering rules were selected. Thus, all results below are for the red_drum_11_75 panel. 

The key output files were:

panel.txt: selected loci and expected heterozygosity

```{r}
panel_txt <- read_table(here::here("data","raw","red_drum_11_75panel.txt"),
                        col_types = cols(Pos = col_character()))
```

```{r}
panel_txt
```

Minimum expected heterozygosity (He) across WGULF and EGULF

```{r}
panel_txt %>%
  summarize(mean_he = mean(summaryHe),
            min_he = min(summaryHe),
            max_he = max(summaryHe),
            n_loci = n())
```

primers.txt: forward and reverse primer sequences

```{r}
primers_txt <- read_table(here::here("data","raw","red_drum_11_75primers.txt"), 
                                                  col_types = cols(Pos = col_character()))
```

```{r}
primers_txt
```

pos.txt: SNP positions relative to the amplicon

```{r}
pos_txt <- read_table(here::here("data","raw","red_drum_11_75pos.txt"), 
                                                  col_types = cols(Pos = col_character()))
```

```{r}
pos_txt
```

fwd.txt: forward primer list for downstream genotyping

```{r}
fwd_txt <- read_table(here::here("data","raw","red_drum_11_75fwd.txt"), 
                                                  col_types = cols(Pos = col_character()))
```

```{r}
fwd_txt
```

ampRef.fa: reference amplicon sequences

# Sanity check of amplicons

We expect the 75 bp design corresponds to:

50 bp microhap windows
variants targeted within the first 75 bp of the amplicon
variants throughout the genome

```{r}
check <- panel_txt %>%
  select(-Chr,-Pos) %>%
  left_join(primers_txt, by = "Locus") %>%
  select(Locus, wStart, wEnd,orient, genomeStart,Pos, Chr) %>%
  mutate(micr_window = wEnd - wStart + 1) %>%
  mutate(last_variant_pos_on_amplicon = case_when(
    orient == "left" ~ wEnd - genomeStart + 1))
```

```{r}
ggplot(check, aes(x = micr_window)) + 
  geom_histogram()


ggplot(check, aes(x = last_variant_pos_on_amplicon)) + 
  geom_histogram()

ggplot(panel_txt, aes(x = Pos)) +
  geom_rug() +
  facet_wrap(~Chr, scales = "free_x") +
  theme_bw()
```

```{r}
ggplot(panel_txt, aes(x = wStart, y = 1)) +
  geom_point(size = 2) +
  facet_wrap(~Chr, scales = "free_x") +
  theme_bw() +
  labs(x = "Genomic position", y = NULL) + 
  theme(axis.text.y = element_blank())
```

Looks good!

# Panel summary

The makeup of the target amplicons were summarized with an R script (summarize_panel_runs.R)

```{r}
panel_summ <- read_tsv(here::here("data","derived","panel_run_summary.tsv"))
```

```{r}
knitr::kable(
  panel_summ %>% select(-run),
  digits = 2,
  caption = "Summary of 500 locus panel"
)
```

Last, do that primer program that Ali did and you did for the oysters

# 

