---
title: "soce_dzyn26_initial_primer_design"
format: html

execute:
  warning: false
  message: false
---

```{r, include = FALSE}
library(tidyverse)
library(readxl)
library(here)
```

I followed Tom Delomas' primer design pipeline to generate 500 primers for a new GT-seq panel for red drum.

# Raw reads

Chris Hollenbeck provided whole genome sequence data. They are located on the HPC in /marinegenomics/projects/fish/red_drum/drum_pop_wgs/raw/. Below is a quick overview of their metadata:

```{r}
meta_data <- readxl::read_xlsx(here::here("data","raw","drum_wgs_samples.xlsx"))
```

```{r}
knitr::kable(
  meta_data,
  digits = 2,
  caption = "Metadata of WGS Red Drum Samples"
)
```

# Genotyping

The raw reads were genotyped using Eriq Anderson's mega-non-model-wgs-snakeflow pipeline https://github.com/eriqande/mega-non-model-wgs-snakeflow

There were several adjustments that were required before the pipeline could function normally. Details of adjustments made can be found in another document, "mega_snakeflow_on_crest."

QC of the genotyping can be found here:

[View QC results](multiqc.html)

The working SNP set used for primer design was made from manually extracting biallelic snps from the pass-maf-0.05 bcf file generated from the pipeline.

The pass-maf-0.05 file was generated by these filters in GATK:

QD \< 2.0 filtered QUAL \< 30.0 filtered SOR \> 3.0 filtered FS \> 60.0 filtered MQ \< 40.0 filtered MQRankSum \< -12.5 filtered ReadPosRankSum \< -8.0 filtered

and having a MAF of 0.05 or greater.

The working set of SNPs was called red_drum_maf005.snps.vcf.gz and was generated by manually selected biallelic snps with: bcftools view -v snps -m2 -M2

# Phasing with WhatsHap

Each sample was phased independently using WhatsHap. The phased VCFs were generated from the filtered SNP VCF and corresponding overlap-clipped BAM files.

A representative command was:

whatshap phase\
--reference /work/marinegenomics/jmatt1/wgs/mega_soce_cmh/resources/genome.fasta\
--sample SAMPLE_NAME\
--output SAMPLE.whatshap.vcf.gz\
red_drum.maf005.snps.vcf.gz\
overlap_clipped/SAMPLE.bam

Each phased VCF was indexed, and all 20 single-sample phased VCFs were merged into a single multi-sample phased VCF:

bcftools merge single_sample_vcfs/\*.whatshap.vcf.gz\
-Oz -o allWhatshap_red_drum_maf005.vcf.gz

bcftools index -t allWhatshap_red_drum_maf005.vcf.gz

The merged VCF contained 20 samples and phased genotype information.

# Population-specific VCFs

The merged phased VCF was split into WGULF and EGULF population-specific VCFs using sample lists.

bcftools view -S WGULF.samples.txt\
allWhatshap_red_drum_maf005.vcf.gz\
-Oz -o allWhatshap_WGULF.vcf.gz

bcftools index -t allWhatshap_WGULF.vcf.gz

bcftools view -S EGULF.samples.txt\
allWhatshap_red_drum_maf005.vcf.gz\
-Oz -o allWhatshap_EGULF.vcf.gz

bcftools index -t allWhatshap_EGULF.vcf.gz

Each population-specific VCF contained 10 samples.

# Optional HWE screening

To mimic the oyster-panel workflow, Hardy-Weinberg Equilibrium filtering was evaluated separately within WGULF and EGULF using PLINK.

VCF variant IDs were first set using chromosome and position:

bcftools annotate --set-id '%CHROM:%POS' allWhatshap_EGULF.vcf.gz -Oz -o allWhatshap_EGULF.ids.vcf.gz bcftools annotate --set-id '%CHROM:%POS' allWhatshap_WGULF.vcf.gz -Oz -o allWhatshap_WGULF.ids.vcf.gz

PLINK HWE tests were run with:

plink --vcf allWhatshap_EGULF.ids.vcf.gz --double-id --allow-extra-chr --hardy --out EGULF_ids plink --vcf allWhatshap_WGULF.ids.vcf.gz --double-id --allow-extra-chr --hardy --out WGULF_ids

SNPs with HWE p-value \< 0.01 in either population were counted. A total of 70,715 SNPs failed this criterion, representing approximately 1.26% of SNPs tested. As a final step, candidate primers were filtered if they included these SNPs.

# Microhaplotype expected heterozygosity

Microhaplotype expected heterozygosity was calculated using CalcHe_mh_vcf.jar.

For each population VCF, two window sizes were evaluated:

50 bp windows, corresponding to the 75 bp sequencing-read design 125 bp windows, corresponding to the 150 bp sequencing-read design

Example commands:

java -jar CalcHe_mh_vcf.jar -v allWhatshap_WGULF.vcf.gz -w 50 mv vcf_he_output.txt He_50_WGULF.txt

java -jar CalcHe_mh_vcf.jar -v allWhatshap_WGULF.vcf.gz -w 125 mv vcf_he_output.txt He_125_WGULF.txt

The same commands were run for EGULF, producing:

He_50_WGULF.txt He_125_WGULF.txt He_50_EGULF.txt He_125_EGULF.txt

# Panel design inputs

Panel design was performed in R using a modified version of the oyster microhaplotype panel-design script.

The population input section was updated to:

inputfiles \<- tibble( pop = c("WGULF", "EGULF") )

For 75 bp read designs, the script used the 50 bp He files:

He_50_WGULF.txt He_50_EGULF.txt

For 150 bp read designs, the script used the 125 bp He files:

He_125_WGULF.txt He_125_EGULF.txt

Expected heterozygosity was summarized across populations using the minimum value:

summaryFunction \<- min

This prioritizes loci that are informative in both WGULF and EGULF.

Other changes of the oyster panel design script were specifying the correct genome:

refGenome \<- "/work/marinegenomics/jmatt1/wgs/mega_soce_cmh/resources/genome.fasta"

and using a Bowtie2 index for red drum.

A Bowtie2 index was built from the same reference genome:

bowtie2-build genome.fasta genome_bt2

and was specified in the panel design R script:

bowtie2_ref \<- "/work/marinegenomics/jmatt1/wgs/mega_soce_cmh/resources/genome_bt2"

Chromosome names and lengths were also added to the R script. These were taken from the ref genome index file.

# Primer design and filtering strategy

Primer design was performed with Primer3 through the panel-design R script.

The Primer3 executable was provided by the microhap conda environment:

primer3path \<- "/home/jmatt1/.conda/envs/microhap/bin/primer3_core"

Primer design was run during each iteration of the greedy selection algorithm. A locus was only selected if valid primers could be designed and the primer pair mapped uniquely to the reference genome using Bowtie2.

The greedy algorithm balanced:

high minimum expected heterozygosity across WGULF and EGULF, even genomic distribution, successful Primer3 primer design, unique primer-pair mapping.

Eight panel-design runs were evaluated:

Rscript panel_design_drum.R red_drum_00_75 0 0 75 Rscript panel_design_drum.R red_drum_10_75 1 0 75 Rscript panel_design_drum.R red_drum_01_75 0 1 75 Rscript panel_design_drum.R red_drum_11_75 1 1 75

Rscript panel_design_drum.R red_drum_00_150 0 0 150 Rscript panel_design_drum.R red_drum_10_150 1 0 150 Rscript panel_design_drum.R red_drum_01_150 0 1 150 Rscript panel_design_drum.R red_drum_11_150 1 1 150

The first numeric argument controlled the no-recombination/complexity filter:

0: allow all loci 1: retain only loci where number of observed haplotypes was no greater than number of SNPs + 1

The second numeric argument controlled adjacent-SNP filtering:

0: allow adjacent SNPs 1: remove loci containing adjacent SNPs

The third numeric argument controlled the intended read-length design:

75: use 50 bp microhap windows 150: use 125 bp microhap windows

# Output files

Each panel-design run produced:

<prefix>panel.txt <prefix>primers.txt <prefix>pos.txt <prefix>fwd.txt <prefix>ampRef.fa

To mimic the mimic the oyster panel design criteria, the 75 bp panels are most appropriate. To target a well-behaving panel, the recombination filter and adjacent SNP-filtering rules were selected. Thus, all results below are for the red_drum_11_75 panel.

The key output files were:

panel.txt: selected loci and expected heterozygosity

```{r}
panel_txt <- read_table(here::here("data","raw","red_drum_11_75panel.txt"),
                        col_types = cols(Pos = col_character()))
```

```{r}
panel_txt
```

Minimum expected heterozygosity (He) across WGULF and EGULF

```{r}
panel_txt %>%
  summarize(mean_he = mean(summaryHe),
            min_he = min(summaryHe),
            max_he = max(summaryHe),
            n_loci = n())
```

primers.txt: forward and reverse primer sequences

```{r}
primers_txt <- read_table(here::here("data","raw","red_drum_11_75primers.txt"), 
                                                  col_types = cols(Pos = col_character()))
```

```{r}
primers_txt
```

pos.txt: SNP positions relative to the amplicon

```{r}
pos_txt <- read_table(here::here("data","raw","red_drum_11_75pos.txt"), 
                                                  col_types = cols(Pos = col_character()))
```

```{r}
pos_txt
```

fwd.txt: forward primer list for downstream genotyping

```{r}
fwd_txt <- read_table(here::here("data","raw","red_drum_11_75fwd.txt"), 
                                                  col_types = cols(Pos = col_character()))
```

```{r}
fwd_txt
```

ampRef.fa: reference amplicon sequences

# Sanity check of amplicons

We expect the 75 bp design corresponds to:

50 bp microhap windows variants targeted within the first 75 bp of the amplicon variants throughout the genome

```{r}
check <- panel_txt %>%
  select(-Chr,-Pos) %>%
  left_join(primers_txt, by = "Locus") %>%
  select(Locus, wStart, wEnd,orient, genomeStart,Pos, Chr) %>%
  mutate(micr_window = wEnd - wStart + 1) %>%
  mutate(last_variant_pos_on_amplicon = case_when(
    orient == "left" ~ wEnd - genomeStart + 1))
```

```{r}
ggplot(check, aes(x = micr_window)) + 
  geom_histogram()


ggplot(check, aes(x = last_variant_pos_on_amplicon)) + 
  geom_histogram()
```

```{r}
ggplot(panel_txt, aes(x = wStart, y = 1)) +
  geom_point(size = 2) +
  facet_wrap(~Chr, scales = "free_x") +
  theme_bw() +
  labs(x = "Genomic position", y = NULL) + 
  theme(axis.text.y = element_blank())
```

Looks good!

# Initial panel summary

The makeup of the target amplicons were summarized with an R script (summarize_panel_runs.R)

```{r}
panel_summ <- read_tsv(here::here("data","derived","panel_run_summary.tsv"))
```

```{r}
knitr::kable(
  panel_summ %>% select(-run),
  digits = 2,
  caption = "Summary of 500 locus panel"
)
```

Note, the amplicon size here does not include the Illumina sequencing tags that are added during PCR1 or the tags that are added during PCR2:

With the sequencing tags:+ 67
With the P5 and i5: + 65

Expected minimum: 79+132 = 211
Expected median: 136+132 = 268
Expected max: 174+132 = 316

```{r, include = FALSE}
# Rename loci here

locus_prefix <- primers_txt %>%
  select(Locus) %>%
  unique() %>%
  mutate(locus_prefix = paste0("soce_",sprintf("%04d",row_number())))

locus_name_key <- primers_txt %>%
  select(Locus, orient) %>%
  left_join(locus_prefix, by = "Locus") %>%
  mutate(locus_id = case_when(
    orient == "left" ~ paste0(locus_prefix, "_F"), 
    TRUE ~ paste0(locus_prefix, "_R")
  )) %>%
  rename(original_prefix = Locus) %>%
  mutate(original_name = paste0(original_prefix,orient)) %>%
  select(-orient)

locus_name_key
```

# Check for interactions

## Run primers through Primer Pooler

```{r, include = FALSE}
primer_seqs <- primers_txt %>%
  mutate(original_name = paste0(Locus, orient)) %>%
  left_join(locus_name_key, by = "original_name") %>%
  select(locus_id, seq)
```

```{r, include = FALSE}
# example of format needed for primer pooler

primer_pool_example <- readRDS("/home/shared/cvir/analysis/cvir_gtseq1_primers/data/derived/primer_pooler")

primer_pool_example
```

```{r, include = FALSE}
primer_pooler <- primer_seqs %>%
  mutate(seq_name = str_replace_all(locus_id, "_", "-")) %>%
  select(seq_name, seq)
```

```{r, include = FALSE}

#Write data file for Primer Pooler to csv
write.csv(primer_pooler, file = here::here("data","derived","primer_pooler.csv"),row.names=FALSE)

#Run script format_primer_pooler.sh in ./bash to make file
```

Settings for primer pooler:

annealing temp is the minimum temp used in your primer design (in our case, 57C) magnesium is 3mM (specific to the Qiagen mastermix we use for gtseq) Sodium (use default setting of 50) dNTP is 0.8mM (specific to Qiagen mastermix we use)

```{r, include = FALSE}
lines <- readLines(here::here("data","derived","primer_pooler_4thresh.txt"))

hits <- tibble(line = lines) %>%
  mutate(row = row_number()) %>%
  filter(str_detect(line, " versus |dG =")) %>%
  mutate(hit_id = cumsum(str_detect(line, " versus "))) %>%
  group_by(hit_id) %>%
  summarize(
    pair = line[str_detect(line, " versus ")][1],
    dG = line[str_detect(line, "dG =")][1],
    .groups = "drop"
  ) %>%
  separate(pair, into = c("primer1", "primer2"), sep = " versus ") %>%
  mutate(
    dG = as.numeric(str_remove(dG, "dG = "))
  ) %>%
  arrange(dG)
```

Interactions identified by primer pooler with dG less than -4

```{r}
hits_loci <- hits %>%
  #mutate(primer1 = str_replace_all(primer1,"-","_")) %>%
  #mutate(primer2 = str_replace_all(primer2, "-","_")) %>%
  mutate(
    locus1 = str_remove(primer1, "-[FR]$"),
    locus2 = str_remove(primer2, "-[FR]$")
  ) %>%
  mutate(locus1 = str_replace_all(locus1,"-","_")) %>%
  mutate(locus2 = str_replace_all(locus2, "-","_"))

hits_loci
```

Going to filter out interactions with a dG less than -5.5 (lower dG, more chance for interaction)

Run a script to remove the fewest number of loci that overcomes the interaction issues

```{r}
edges <- hits_loci %>%
  filter(dG < -5.5) %>%
  select(locus1, locus2)

removed <- character()

while(nrow(edges) > 0){

  counts <-
    c(edges$locus1, edges$locus2) |>
    table() |>
    sort(decreasing = TRUE)

  remove <- names(counts)[1]

  removed <- c(removed, remove)

  edges <-
    edges %>%
    filter(locus1 != remove,
           locus2 != remove)
}

removed
```

Ok, let's remove these:

```{r}
primer_seqs_filt <- primer_seqs %>%
  mutate(locus = str_remove(locus_id, "_[FR]$")) %>%
  filter(!locus %in% removed)
```

Loci remaining:

```{r}
nrow(primer_seqs_filt)/2
```

```{r}
primer_pooler_filt <- primer_seqs_filt %>%
  mutate(seq_name = str_replace_all(locus_id, "_", "-")) %>%
  select(seq_name, seq)
```

Running primer pooler again to confirm:

```{r, include = FALSE}
write.csv(primer_pooler_filt, file = here::here("data","derived","primer_pooler_filt.csv"),row.names=FALSE)
```

```{r,include = FALSE}
lines <- readLines(here::here("data","derived","primer_pooler_filt_thresh4.txt"))

hits <- tibble(line = lines) %>%
  mutate(row = row_number()) %>%
  filter(str_detect(line, " versus |dG =")) %>%
  mutate(hit_id = cumsum(str_detect(line, " versus "))) %>%
  group_by(hit_id) %>%
  summarize(
    pair = line[str_detect(line, " versus ")][1],
    dG = line[str_detect(line, "dG =")][1],
    .groups = "drop"
  ) %>%
  separate(pair, into = c("primer1", "primer2"), sep = " versus ") %>%
  mutate(
    dG = as.numeric(str_remove(dG, "dG = "))
  ) %>%
  arrange(dG)
```

```{r, include = FALSE}
hits_loci <- hits %>%
  #mutate(primer1 = str_replace_all(primer1,"-","_")) %>%
  #mutate(primer2 = str_replace_all(primer2, "-","_")) %>%
  mutate(
    locus1 = str_remove(primer1, "-[FR]$"),
    locus2 = str_remove(primer2, "-[FR]$")
  ) %>%
  mutate(locus1 = str_replace_all(locus1,"-","_")) %>%
  mutate(locus2 = str_replace_all(locus2, "-","_"))

```

```{r}
edges <- hits_loci %>%
  filter(dG < -5.5) %>%
  select(locus1, locus2)

removed <- character()

while(nrow(edges) > 0){

  counts <-
    c(edges$locus1, edges$locus2) |>
    table() |>
    sort(decreasing = TRUE)

  remove <- names(counts)[1]

  removed <- c(removed, remove)

  edges <-
    edges %>%
    filter(locus1 != remove,
           locus2 != remove)
}

removed
```

Confirms that we removed all interactions with dG less than -5.5

## Run blast for final check

Ok, now let's run blast as a final check

```{r}
library(seqhelp) #notice this is version 0.0.0.9000
library(janitor)

CONDA_ENV <- "paneldesign"
CONDA_PATH <- Sys.getenv("CONDA_PATH")
```

self_blast function in seqhelp not operating right, so using slightly modified function

```{r}
self_blast_fixed <- function(seq, id, conda_env = CONDA_ENV, conda_sh_path = CONDA_PATH) {
  tmp_dir <- tempdir()
  tmp_fasta <- glue::glue(tmp_dir, "tmp.fasta", .sep = "/")
  tmp_db <- glue::glue(tmp_dir, "tmp", .sep = "/")
  tmp_out <- glue::glue(tmp_dir, "tmp_out.blast", .sep = "/")

  seqhelp::write_fasta(file = tmp_fasta, seq = seq, names = id)

  invisible({
    seqhelp::systemc(
      glue::glue("makeblastdb -in {tmp_fasta} -dbtype nucl -out {tmp_db}"),
      env = conda_env,
      ignore_stdout = TRUE,
      ignore_stderr = TRUE,
      conda_sh_path = conda_sh_path
    )

    seqhelp::systemc(
      glue::glue("blastn -query {tmp_fasta} -db {tmp_db} -outfmt 6 -task blastn-short > {tmp_out}"),
      env = conda_env,
      conda_sh_path = conda_sh_path
    )
  })

  blast_results <- seqhelp::read_blast(tmp_out)

  blast_results <- blast_results %>%
    dplyr::filter(.data$qseqid != .data$sseqid) %>%
    dplyr::arrange(dplyr::desc(.data$bitscore))

  return(blast_results)
}
```

```{r}
blast_res_filt <- self_blast_fixed(seq = primer_seqs_filt$seq, id = primer_seqs_filt$locus_id)
```

Any primers that are exactly the same?

```{r}
primer_lengths <- primer_seqs %>%
  transmute(
    primer_id = locus_id,
    primer_len = nchar(seq)
  )

identical_flags <- blast_res_filt %>%
  left_join(primer_lengths, by = c("qseqid" = "primer_id")) %>%
  rename(q_len = primer_len) %>%
  left_join(primer_lengths, by = c("sseqid" = "primer_id")) %>%
  rename(s_len = primer_len) %>%
  filter(
    pident == 100,
    length >= 15,
    qstart == 1,
    qend == q_len,
    sstart == 1,
    send == s_len
  ) %>%
  mutate(
    q_locus = str_remove(qseqid, "_[FR]$"),
    s_locus = str_remove(sseqid, "_[FR]$")
  ) %>%
  filter(q_locus != s_locus)

identical_flags
```

```{r}
primer_seqs_filt %>%
  filter(locus_id %in% c("soce_0251_R","soce_0494_R"))
```

Yes, these two are exactly the same

Is it better to remove 251 or 494?

```{r}
hits_loci %>%
  filter(grepl("0251",locus1) | grepl("0251",locus2))

hits_loci %>%
  filter(grepl("0494",locus1) | grepl("0494",locus2))
```

Better to remove 251.

```{r}
filtered_primers <- primer_seqs_filt %>%
  filter(!grepl("0251",locus))
```

I did not use the blast results to filter out any other loci, but am open to it.

# Check if any SNPs failing HWE are in the filtered loci set

```{r}
hwe_fail <- readLines(here::here("data","derived","HWE_fail_any_pop.sites"))

panel_txt_filt <- panel_txt %>%
  rename(original_prefix = Locus) %>%
  left_join(locus_name_key, by = "original_prefix") %>%
  filter(locus_prefix %in% filtered_primers$locus) %>%
  select(-locus_id, -original_prefix, -original_name) %>%
  distinct()

panel_snps <- panel_txt_filt %>%
  separate_rows(Pos, sep = ",") %>%
  mutate(
    snp_id = paste0(Chr, ":", Pos)
  )
```

```{r}
panel_hwe_hits <- panel_snps %>%
  filter(snp_id %in% hwe_fail)

panel_hwe_hits %>%
  count(locus_prefix, name = "n_failed_snps") %>%
  arrange(desc(n_failed_snps))

```

Yes, looks like there are 18 amplicons that would contain at least one of these snps. Let's filter these out.

```{r}
remove_hwe <- panel_hwe_hits %>%
  count(locus_prefix, name = "n_failed_snps") %>%
  arrange(desc(n_failed_snps)) %>%
  pull(locus_prefix)
```

```{r}
final_set_primers <- filtered_primers %>%
    filter(!locus %in% remove_hwe)

final_set_primers
```

417 primer combinations (F & R) remaining

# Overview of final set of primers

```{r, include = FALSE}
panel_txt_final <- panel_txt %>%
  rename(original_prefix = Locus) %>%
  left_join(locus_name_key, by = "original_prefix") %>%
  filter(locus_prefix %in% final_set_primers$locus) %>%
  select(-locus_id, -original_prefix, -original_name) %>%
  distinct()
```

```{r}
ggplot(panel_txt_final , aes(x = wStart, y = 1)) +
  geom_point(size = 2) +
  facet_wrap(~Chr, scales = "free_x") +
  theme_bw() +
  labs(x = "Genomic position", y = NULL) + 
  theme(axis.text.y = element_blank())
```

Interesting, what is going on with CP136871.1...